Algorithms for the Automated Correction of Vertical Drift in Eye Tracking Data

### 2020-10-23 manuscript/manuscript.tex
Revisions following round 1 of reviews

\documentclass[doc,biblatex]{apa7}
\DeclareLanguageMapping{american}{american-apa}
\addbibresource{refs.bib}
\usepackage{amssymb} % For producing checkmark character
\captionsetup[figure]{textfont=rm,font=small} % Override italic APA figure captions
\newcommand\possessivecite[1]{\citeauthor{#1}'s (\citeyear{#1})} % Possessive citation command

@@

\abstract{A common problem in eye tracking research is vertical drift---the progressive displacement of fixation registrations on the vertical axis that results from a gradual loss of eye tracker calibration over time. This is particularly problematic in experiments that involve the reading of multiline passages, where it is critical that fixations on one line are not erroneously recorded on an adjacent line. Correction is often performed manually by the researcher, but this process is tedious, time-consuming, and prone to error and inconsistency. Various methods have previously been proposed for the automated, post-hoc correction of vertical drift in reading data, but these methods vary greatly, not just in terms of the algorithmic principles on which they are based, but also in terms of their availability, documentation, implementation languages, and so forth. Furthermore, these methods have largely been developed in isolation with little attempt to systematically evaluate them, meaning that drift correction techniques are moving forward blindly. We document nine major algorithms, including two that are novel to this paper, and evaluate them using both simulated and natural eye tracking data. Our results suggest that a method based on dynamic time warping offers great promise, but we also find that some algorithms are better suited than others to particular types of drift phenomena and reading behavior, allowing us to offer evidence-based advice on algorithm selection.}

@@

Reading is a fundamental skill for navigating modern society and, as such, is subject to intense study in the cognitive and language sciences. Among the many tools that researchers use to investigate reading in the laboratory, eye tracking occupies a prominent position. Using this technique, participants' eye movements may be recorded as they read written material, providing a window into the relevant cognitive processes as they unfold. Technological advancements in eye tracking, particularly from the 1970s \parencite[see e.g.,][]{Rayner:1998}, have allowed scholars to collect increasingly accurate measures of eye movements during reading tasks, leading to great improvements in the investigation of the cognitive processes underlying reading and reading acquisition.

@@

In our everyday experience, however, we often do not encounter sentences in isolation; a good part of our reading experience involves connected text that is distributed over multiple lines. Therefore, experiments based on paragraph reading also provide insight into the reading experience, while allowing us to address levels of processing that are simply not available when one reads a single sentence, such as the role of broader context or the integration of syntactic relations across sentence boundaries \parencite{Jarodzka:2017}. Indeed, studies of multiline reading have become more prevalent in recent years, with researchers using passage reading tasks to investigate, for example, the effect of text- and participant-level characteristics on eye movements \parencite{Kuperman:2018} or of contextual facilitation on developing readers' eye movements \parencite{TiffinRichards:2020}. Eye tracking databases of multiline reading, such as GECO \parencite{Cop:2017} and Provo \parencite{Luke:2018}, have also been made available.

A technical issue that arises from the particular circumstances of multiline reading is so-called ``vertical drift,'' which we define as the progressive displacement of fixation registrations on the vertical axis \textit{over time}. In other words, fixations may be recorded above or below the line of text that the participant was actually reading, and the amount of measurement error may fluctuate over time, making it nontrivial to eliminate. Fig.~\ref{fig01}A depicts a reading trial exhibiting vertical drift phenomena; in this case, fixations---especially those on the left hand side---are recorded around one line higher than where the reader was actually fixating, but they also tend to slope down to the right such that fixations on the right hand side seem to be better aligned.

Vertical drift can occur quite unpredictably, even following good quality calibration, and it is likely caused by subtle movements in head position or changes in pupil dilation, which can be difficult to control for, even in a laboratory setting. Such sources of measurement error are often exacerbated in the context of multiline reading because the reading of a passage of text necessarily takes longer than single words or sentences, during which time calibration may begin to degrade. In addition, there are less frequent opportunities to recalibrate the device, since calibration can only be performed between trials and not during the reading of a passage.

@@

A number of methods have previously been developed to automate post-hoc vertical drift correction. \texttt{FixAlign}, an R package developed by \textcite{Cohen:2013}, is currently the most well-established method in the experimental psychology community, although other methods have recently been proposed by \textcite{Schroeder:0000} and \textcite{Spakov:2019}. In addition, there is a disparate body of work from several subfields of computer science, such as biometrics \parencite{Abdulin:2015}, educational technology \parencite{Hyrskykari:2006}, and user-interface design \parencite{Beymer:2005}, in which various ad-hoc algorithms have been reported \parencite[see also][]{LimaSanches:2015,Carl:2013,Mishra:2012,MartinezGomez:2012,Yamaya:2017}.

These reported methods can be difficult to evaluate and use because they vary widely in terms of their availability, design choices, implementation languages, usability, level of documentation, expected input data, and the extent to which they rely on project-specific heuristics or particular eye tracker hardware. Furthermore, these methods have largely been developed in isolation from each other, and there has been little attempt to systematically evaluate them, so drift correction software is moving forward blindly without an evidence base to support new directions. In this paper, we attempt to classify the reported methods into nine major approaches, which we formalize as nine simple algorithms that adopt a consistent design model. In other words, we do not attempt to evaluate existing software implementations; rather, we explore the spectrum of drift correction algorithms by isolating and evaluating the core principles on which previous methods have been based. Our goal is to provide a systematic comparison of these algorithms in order to guide researchers' choices about the most suitable methods and to lay a solid foundation for future drift correction software.

To be clear, the algorithms we consider in this paper are restricted to one specific type of problem. Firstly, we only consider algorithms intended for use on passages of text; other uses of eye tracking, such as visual search and scene perception, can also undergo drift correction, but the methods required are different \parencite[see e.g.,][]{Zhang:2011,Zhang:2014}. Secondly, we only consider the problem of post-hoc correction; vertical drift can also be corrected in real time, but this imposes a more restrictive set of constraints that are better handled by other types of algorithm \parencite[see e.g.,][]{Hyrskykari:2006,Palmer:2016}. Thirdly, we only consider fully automated algorithms that do not require human supervision; an alternative approach would make greater use of human insight.

@@

In this section, we describe nine algorithms for the automated, post-hoc correction of vertical drift. The reader may also wish to refer to the Appendix where the algorithms are presented in pseudocode alongside other technical details.

@@

The \texttt{attach} algorithm is the simplest of the algorithms considered in this paper. The algorithm simply attaches each fixation to its closest line, as illustrated in Fig.~\ref{fig02}A. While this has the benefit of being extremely simple, it is generally not resilient to the kinds of drift phenomena described above. However, \texttt{attach} serves as a useful baseline algorithm, since it essentially corresponds to an eye tracking analysis in which no correction was performed---a standard analysis of eye tracking data would simply map fixations to the closest words or other areas of interest. We return to this point later in the paper.

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/fig02_double_column.eps}}
	\vspace*{2pt}
	\caption{Illustrations of the core principle of each algorithm. The colors represent which line each fixation was assigned to (black indicates line one, etc.). \texttt{attach} maps the fixation on ``sit'' to line 2, while all other algorithm correctly map it to line 1.}
	\label{fig02}
	\end{figure}

@@

The \texttt{chain} algorithm is based closely on one of the methods implemented in the R package \texttt{popEye} \parencite{Schroeder:0000} and can be seen as an extension of \texttt{attach}. Fixations are first linked together into ``chains''---sequences of consecutive fixations that are within a specified \textit{x}~and \textit{y}~distance of each other (see Fig.~\ref{fig02}B for an example). Fixations within a chain are then attached to whichever line is closest to the mean of their \textit{y}~values. This procedure is similar to the methods reported by \textcite{Hyrskykari:2006} and \textcite{Mishra:2012}, so we consider these to be special cases of \texttt{chain}.

The \texttt{chain} algorithm generally provides better performance over \texttt{attach} by exploiting the sequence's order information, as illustrated in Fig.~\ref{fig02}B: The fixation on ``sit'' on line~1 is chained to two surrounding fixations and is correctly mapped to line~1; while \texttt{attach} assigns this fixation to line~2 to which it is technically closer. A disadvantage of the method, however, is that it is necessary to specify appropriate thresholds that determine when a new chain begins. If these thresholds are set too low, \texttt{chain} becomes equivalent to \texttt{attach}; if they are set too high, \texttt{chain} will group large numbers of fixations together and force them onto a single inappropriate line. By default, \texttt{popEye} sets the \textit{x}~threshold to 20 $\times$ the font height and the \textit{y}~threshold to 2 $\times$ the font height. It is not exactly clear how these defaults were chosen, but we would tentatively suggest that the \textit{x}~threshold should be set to approximately one long saccade length (we use 192~px), and the \textit{y}~threshold to around half a line height (we use 32~px).

@@

The \texttt{cluster} algorithm is also based on one of the methods implemented in \texttt{popEye} \parencite{Schroeder:0000}. \texttt{cluster} applies \textit{k}-means clustering\footnote{We used the \texttt{KMeans} function from the Python library \texttt{scikit-learn} \parencite{scikit-learn}. There are many variations of \textit{k}-means clustering, which we have not systematically compared.} to the \textit{y}~values of all fixations in order to group the fixations into $m$ clusters, where $m$ is the number of lines in the passage. Once each fixation has been assigned to a cluster, clusters are mapped to lines based on the mean \textit{y}~values of their constituent fixations: The cluster with the smallest mean \textit{y}~value is assigned to line one and so forth.

Unlike \texttt{attach} and \texttt{chain}, \texttt{cluster} does not assign fixations to the closest line in absolute terms; instead, it operates on the principle that fixations with similar \textit{y}~values must belong to the same line regardless of how far away that line might be. As such, the algorithm generally handles drift issues quite well. However, \texttt{cluster} will often not perform well if there is even quite mild overlap between fixations from different lines. In addition, since \textit{k}-means clustering is not guaranteed to converge on the same set of clusters on every run, the \texttt{cluster} algorithm is nondeterministic and can be somewhat unpredictable across multiple runs on the same reading trial. This is an important consideration from the point of view of reproducible research output.

@@

The \texttt{compare} algorithm is directly based on the method reported by \textcite{LimaSanches:2015} and is very similar to the more complex methods described by \textcite{Yamaya:2017}. The fixation sequence is first segmented into ``gaze lines'' by identifying the return sweeps---long saccades that move the eye from the end of one line to the start of the next. The algorithm considers any saccade that moves from right to left by more than some threshold value (we use 512~px) to be a return sweep. Gaze lines are then matched to text lines based on a measure of similarity between them. \textcite{LimaSanches:2015} considered three measures of similarity and found dynamic time warping \parencite{Vintsyuk:1968,Sakoe:1978} to be the best method. Similarly, \textcite{Yamaya:2017} use the closely related Needleman--Wunsch algorithm \parencite{Needleman:1970}. In the example in Fig.~\ref{fig02}D, we only depict the first gaze line (black dots), which is compared to each line of text; the best match is with the first text line, as reflected in its low cost (i.e., the sum of the mapping lines is minimized).

The gaze lines and text lines are only compared in terms of their \textit{x}~values under the assumption that the fixations in a gaze line should have a good horizontal alignment with the centers of the words in the corresponding text line. Relying only on the \textit{x}~values helps the algorithm overcome vertical drift issues, but it is also problematic because the lines of text in a passage tend to be horizontally similar to each other; each line tends to contain a similar number of words that are of a similar length, resulting in potential ambiguity about how gaze lines and text lines should be matched up. To alleviate this issue, both \textcite{LimaSanches:2015} and \textcite{Yamaya:2017} only compare the gaze line to a certain number of nearby text lines (we set this parameter to 3, which is effectively the closest line plus one line above and one line below).

@@

The \texttt{merge} algorithm is closely based on the post-hoc correction method described by \textcite{Spakov:2019}. The algorithm begins by creating ``progressive sequences''---consecutive fixations that are sufficiently close together. This is similar to \texttt{chain}, except that the sequences are strictly progressive (they only move forward), so a regression will initiate a new progressive sequence. The original method uses several parameters to define what constitutes ``sufficiently close together,'' but here we simplify this down to a single parameter, the \texttt{y\_thresh} which determines how close the \textit{y}~values of two consecutive fixations must be to be considered part of the same progressive sequence (we use 32~px).

Once these sequences have been created, they are repeatedly merged into larger and larger sequences until the number of sequences is reduced to $m$---one for each line of text. On each iteration of the merge process, the algorithm fits a regression line to every possible pair of sequences (with the proviso that the two sequences must contain some minimum number of fixations). If the absolute gradient of the regression line or its error (root-mean-square deviation) is above a threshold (we use 0.1 and 20 respectively), the candidate merger is abandoned. The intuition here is that, if two sequences belong to the same text line, the regression line fit to their combined fixations will have a gradient close to 0 and low regression error. Of the candidate mergers that remain, the pair of sequences with the lowest error are merged and added to the pool of sequences, replacing the original two sequences and reducing their number by one. This process is repeated until no further mergers are possible.

@@

The \texttt{regress} algorithm, which is closely based on \possessivecite{Cohen:2013} R package \texttt{FixAlign}, treats the fixations as a cloud of unordered points and fits $m$ regression lines to this cloud. These regression lines are parameterized by a slope, vertical offset, and standard deviation, and the best parameters are obtained by minimizing\footnote{We used the \texttt{minimize} function from the Python library \texttt{SciPy} \parencite{Virtanen:2020}. We have not systematically compared the choice of optimizer settings.} an objective function that determines the overall fit of the lines through the fixations. The algorithm has six free parameters which are used to specify the lower and upper bounds of the slope, offset, and standard deviation. Here we directly adopt \texttt{FixAlign}'s defaults: $[-0.1, 0.1]$, $[-50, 50]$, and $[1, 20]$ respectively. Once the $m$ best-fitting regression lines are obtained (e.g., the dashed lines in Fig.~\ref{fig02}F), \texttt{regress} assigns each fixation to the highest-likelihood regression line, which in~turn is associated with a text line.

@@

As far as we know, the \texttt{split} algorithm takes an approach that is distinct from anything previously reported, although it is conceptually similar to \texttt{segment}. Like \texttt{segment}, the \texttt{split} algorithm works on the principle of splitting the fixation sequence into subsequences by first identifying the return sweeps. However, \texttt{split} is not restricted to finding exactly $m-1$ return sweeps; instead, it identifies the most likely set of return sweeps, however many that turns out to be. There are various ways of approaching this classification problem, but here we use \textit{k}-means clustering to partition the set of saccades into exactly two clusters. Since return sweeps are usually highly divergent from normal saccades, one of the two clusters will invariably contain the return sweeps, which can then be used to split the fixation sequence into subsequences. However, since this is not guaranteed to produce $m-1$ return sweeps (and therefore $m$ subsequences) an order-based mapping is not possible, so \texttt{split} must use absolute position: Subsequences are mapped to the closest text lines in absolute terms. \texttt{split} has the advantage of generally finding all true return sweeps, and even if it identifies some false positives, the resulting subsequences can still be mapped to the appropriate lines by absolute position. However, this also means the algorithm is less resilient to vertical drift issues.

@@

The final algorithm we consider, \texttt{warp}, is also novel to this paper, although it is mostly a wrapper around a preexisting algorithm---dynamic time warping \parencite[DTW;][]{Vintsyuk:1968,Sakoe:1978}. DTW was used by the \texttt{compare} algorithm to provide a measure of dissimilarity between a gaze line and a text line. To our knowledge, however, there have been no previous reports of DTW being used for its primary purpose: finding the best alignment between two time sequences. This is somewhat surprising because DTW is the natural computational choice for tackling vertical drift. Two methods have previously been described that share some conceptual similarities. \textcite{Carl:2013} uses a basket of reading-related measures to place a cost on different paths through a lattice of fixation-to-character mappings and selects the path with minimal cost. \textcite{MartinezGomez:2012} convert the fixation sequence into a low resolution image and find the best transformation of this image to fit the overall shape of the passage. Both of these methods are quite convoluted, however, and we consider them to be special cases of \texttt{warp}, which is a direct application of the standard DTW algorithm to eye tracking data.

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/fig03_single_column.eps}}
	\vspace*{2pt}
	\caption{Illustration of the \texttt{warp} algorithm. The veridical fixation sequence is represented in blue, and the expected fixation sequence (the sequence of word centers) is represented in red. The dashed black lines show the DTW warping path---the optimal way to align the two sequences, such that the sum of the Euclidean distances between matched points (i.e., the sum of the dashed lines) is minimized.}
	\label{fig03}
	\end{figure}

>>>>

\documentclass[doc,biblatex]{apa7}
\DeclareLanguageMapping{american}{american-apa}
\addbibresource{refs.bib}
\usepackage{amssymb} % For producing checkmark character
\usepackage{amsmath} % For making argmin look nice
\captionsetup[figure]{textfont=rm,font=small} % Override italic APA figure captions
\newcommand\possessivecite[1]{\citeauthor{#1}'s (\citeyear{#1})} % Possessive citation command

@@

\abstract{A common problem in eye tracking research is vertical drift---the progressive displacement of fixation registrations on the vertical axis that results from a gradual loss of eye tracker calibration over time. This is particularly problematic in experiments that involve the reading of multiline passages, where it is critical that fixations on one line are not erroneously recorded on an adjacent line. Correction is often performed manually by the researcher, but this process is tedious, time-consuming, and prone to error and inconsistency. Various methods have previously been proposed for the automated, post-hoc correction of vertical drift in reading data, but these methods vary greatly, not just in terms of the algorithmic principles on which they are based, but also in terms of their availability, documentation, implementation languages, and so forth. Furthermore, these methods have largely been developed in isolation with little attempt to systematically evaluate them, meaning that drift correction techniques are moving forward blindly. We document ten major algorithms, including two that are novel to this paper, and evaluate them using both simulated and natural eye tracking data. Our results suggest that a method based on dynamic time warping offers great promise, but we also find that some algorithms are better suited than others to particular types of drift phenomena and reading behavior, allowing us to offer evidence-based advice on algorithm selection.}

@@

Reading is a fundamental skill for navigating modern society and, as such, is subject to intense study in the cognitive and language sciences. Among the many tools that researchers use to investigate reading in the laboratory, eye tracking occupies a prominent position. Using this technique, participants' eye movements may be recorded as they read written material, providing a window into the relevant cognitive processes as they unfold. Technological advancements in eye tracking, particularly from the 1970s \parencite[see e.g.,][]{Rayner:1998}, have allowed researchers to collect increasingly accurate measures of eye movements during reading tasks, leading to great improvements in the investigation of the cognitive processes underlying reading and reading acquisition.

@@

In our everyday experience, however, we often do not encounter sentences in isolation; a good part of our reading experience involves connected text that is distributed over multiple lines. Therefore, experiments based on paragraph reading also provide insight into the reading experience, while allowing us to address levels of processing that are simply not available when one reads a single sentence, such as the role of broader context or the integration of syntactic relations across sentence boundaries \parencite{Jarodzka:2017}. Indeed, studies of multiline reading have become more prevalent in recent years, with researchers using passage reading tasks to investigate, for example, the effect of text- and participant-level characteristics on eye movements \parencite{Kuperman:2018} or of contextual facilitation on developing readers' eye movements \parencite{TiffinRichards:2020}. Several multiline-reading datasets have also been released, including GECO \parencite{Cop:2017}, MECO \parencite{Kuperman:0000}, and Provo \parencite{Luke:2018}.

A technical issue that arises from the particular circumstances of multiline reading is so-called ``vertical drift,'' which we define as the progressive displacement of fixation registrations on the vertical axis \textit{over time}. In other words, fixations may be recorded above or below the line of text that the participant was actually reading, \NEWMATERIAL{and the degree and directionality of this error may fluctuate dynamically with each subsequent fixation}, making it nontrivial to eliminate. Fig.~\ref{fig01}A depicts a reading trial exhibiting vertical drift phenomena; in this case, fixations---especially those on the left hand side---are recorded around one line higher than where the reader was actually fixating, but they also tend to slope down to the right such that fixations on the right hand side seem to be better aligned.

Vertical drift can occur quite unpredictably, even following good quality calibration, \NEWMATERIAL{and it is likely caused by spatial phenomena such as degraded eye tracker calibration at the extremes of the screen, or temporal phenomena such as subtle movements in head position or pupil dilation, which can be difficult to control for, even in a laboratory setting. Such sources of measurement error are often exacerbated in the context of multiline reading because, in comparison to single words or sentences, passages of text are distributed over a larger portion of the screen, including areas where general calibration may be worse, and they take longer to read, during which time calibration may begin to degrade}. There are also less frequent opportunities to recalibrate the device during passage reading, since this can only be performed between trials or pages and not during the reading of a passage.

@@

A number of methods have previously been developed to automate post-hoc vertical drift correction. FixAlign, an R package developed by \textcite{Cohen:2013}, is currently the most well-established method in the experimental psychology community, although other methods have recently been proposed by \textcite{Schroeder:0000} and \textcite{Spakov:2019}. In addition, there is a disparate body of work from several subfields of computer science, such as biometrics \parencite{Abdulin:2015}, educational technology \parencite{Hyrskykari:2006}, and user-interface design \parencite{Beymer:2005}, in which various ad-hoc algorithms have been reported \parencite[see also][]{LimaSanches:2015,Carl:2013,Mishra:2012,MartinezGomez:2012,Yamaya:2017,Palmer:2016,Nussli:2011,Sibert:2000,Lohmeier:2015}.

These reported methods can be difficult to evaluate and use because they vary widely in terms of their availability, design choices, implementation languages, usability, level of documentation, expected input data, and the extent to which they rely on project-specific heuristics or particular eye tracker hardware. Furthermore, these methods have largely been developed in isolation from each other, and there has been little attempt to systematically evaluate them, so drift correction software is moving forward blindly without an evidence base to support new directions. In this paper, we attempt to classify the reported methods into ten major approaches, which we formalize as ten simple algorithms that adopt a consistent design model. In other words, we do not attempt to evaluate existing software implementations; rather, we explore the spectrum of drift correction algorithms by isolating and evaluating the core principles on which previous methods have been based. Our goal is to provide a systematic comparison of these algorithms in order to guide researchers' choices about the most suitable methods and to lay a solid foundation for future drift correction software.

To be clear, the algorithms we consider in this paper are restricted to one specific problem. \NEWMATERIAL{Firstly, we only consider algorithms designed for the ordinary reading of passages of text; other uses of eye tracking, such as visual search and scene perception, can also undergo drift correction, but the methods required are quite different \parencite[see e.g.,][]{Zhang:2011,Zhang:2014}. Similarly, the reading of source code has received some attention, but the affordances and constraints in this domain are quite different from ordinary linguistic reading \parencite[see e.g.,][]{Palmer:2016,Nussli:2011}.} Secondly, we only consider the problem of post-hoc correction; vertical drift can also be corrected in real time, but this imposes a more restrictive set of constraints that are better handled by other types of algorithm \parencite[see e.g.,][]{Hyrskykari:2006,Sibert:2000}. Thirdly, we only consider fully automated algorithms that do not require human supervision.

@@

In this section, we describe ten algorithms for the automated, post-hoc correction of vertical drift. The reader may also wish to refer to Supplementary Item~1 where we present the algorithms in pseudocode alongside other technical details.

@@

The \texttt{attach} algorithm is the simplest of the algorithms considered in this paper. The algorithm simply attaches each fixation to its closest line. While this has the benefit of being extremely simple, it is generally not resilient to the kinds of drift phenomena described above. However, \texttt{attach} serves as a useful baseline algorithm, since it essentially corresponds to an eye tracking analysis in which no correction was performed---a standard analysis of eye tracking data would simply map fixations to the closest words or other areas of interest. We return to this point later in the paper.

@@

The \texttt{chain} algorithm is based closely on one of the methods implemented in the R package popEye \parencite{Schroeder:0000} and can be seen as an extension of \texttt{attach}. Fixations are first linked together into ``chains''---sequences of consecutive fixations that are within a specified \textit{x}~and \textit{y}~distance of each other. Fixations within a chain are then attached to whichever line is closest to the mean of their \textit{y}~values. This procedure is similar to the slightly more complex methods reported by \textcite{Hyrskykari:2006} and \textcite{Mishra:2012}, so we consider these to be special cases of \texttt{chain}.

The \texttt{chain} algorithm generally provides better performance over \texttt{attach} by exploiting the sequence's order information. A disadvantage of the method, however, is that it is necessary to specify appropriate thresholds that determine when a new chain begins. If these thresholds are set too low, \texttt{chain} becomes equivalent to \texttt{attach}; if they are set too high, \texttt{chain} will group large numbers of fixations together and force them onto a single inappropriate line. By default, popEye sets the \textit{x}~threshold to 20 $\times$ the font height and the \textit{y}~threshold to 2 $\times$ the font height. It is not exactly clear how these defaults were chosen, but we would tentatively suggest that the \textit{x}~threshold should be set to approximately one long saccade length (we use 192~px), and the \textit{y}~threshold to around half a line height (we use 32~px).

@@

The \texttt{cluster} algorithm is also based on one of the methods implemented in popEye \parencite{Schroeder:0000}. \texttt{cluster} applies \textit{k}-means clustering\footnote{In both the \texttt{cluster} and \texttt{split} algorithms, we used the \texttt{KMeans} function from the Python library Scikit-learn \parencite{scikit-learn}. There are many variations of \textit{k}-means clustering, which we have not systematically compared.} to the \textit{y}~values of all fixations in order to group the fixations into $m$ clusters, where $m$ is the number of lines in the passage. Once each fixation has been assigned to a cluster, clusters are mapped to lines based on the mean \textit{y}~values of their constituent fixations: The cluster with the smallest mean \textit{y}~value is assigned to line one and so forth.

Unlike \texttt{attach} and \texttt{chain}, \texttt{cluster} does not assign fixations to the closest line in absolute terms; instead, it operates on the principle that fixations with similar \textit{y}~values must belong to the same line regardless of how far away that line might be. As such, the algorithm generally handles drift issues quite well. However, \texttt{cluster} will often not perform well if there is even quite mild overlap between fixations from different lines. In addition, since \textit{k}-means clustering is not guaranteed to converge on the same set of clusters on every run, the \texttt{cluster} algorithm is nondeterministic and can be somewhat unpredictable across multiple runs on the same reading trial, which is an important consideration from the point of view of reproducible research output.

@@

The \texttt{compare} algorithm is directly based on the method reported by \textcite{LimaSanches:2015} \NEWMATERIAL{and is very similar to the more complex methods described by \textcite{Yamaya:2017}}. The fixation sequence is first segmented into ``gaze lines'' by identifying the return sweeps---long saccades that move the eye from the end of one line to the start of the next. The algorithm considers any saccade that moves from right to left by more than some threshold value (we use 512~px) to be a return sweep. Gaze lines are then matched to text lines based on a measure of similarity between them. \textcite{LimaSanches:2015} considered three measures of similarity and found dynamic time warping \parencite[DTW;][]{Vintsyuk:1968,Sakoe:1978} to be the best method \NEWMATERIAL{(we discuss DTW in more detail later in this section)}. \NEWMATERIAL{Similarly, \textcite{Yamaya:2017} use the closely related Needleman--Wunsch algorithm \parencite{Needleman:1970}.}

The gaze lines and text lines are compared in terms of their \textit{x}~values under the assumption that the fixations in a gaze line should have a good horizontal alignment with the centers of the words in the corresponding text line. Relying only on the \textit{x}~values helps the algorithm overcome vertical drift issues, but it is also problematic because in many standard reading scenarios the lines of text in a passage tend to be horizontally similar to each other; each line tends to contain a similar number of words that are of a similar length, resulting in potential ambiguity about how gaze lines and text lines should be matched up. To alleviate this issue, both \textcite{LimaSanches:2015} \NEWMATERIAL{and \textcite{Yamaya:2017}} only compare the gaze line to a certain number of nearby text lines (we set this parameter to 3, \NEWMATERIAL{which is effectively the closest line plus one line above and one line below}).

@@

The \texttt{merge} algorithm is closely based on the post-hoc correction method described by \textcite{Spakov:2019}. The algorithm begins by creating ``progressive sequences''---consecutive fixations that are sufficiently close together. This is similar to \texttt{chain}, except that the sequences are strictly progressive (they only move forward), so a regression will initiate a new progressive sequence. The original method uses several parameters to define what constitutes ``sufficiently close together,'' but here we boil this down to a single parameter, the \texttt{y\_thresh}, which determines how close the \textit{y}~values of two consecutive fixations must be to be considered part of the same progressive sequence (we use 32~px).

Once these sequences have been created, they are repeatedly merged into larger and larger sequences until the number of sequences is reduced to $m$, one for each line of text. On each iteration of the merge process, the algorithm fits a regression line to every possible pair of sequences (with the proviso that the two sequences must contain some minimum number of fixations). If the absolute gradient of the regression line or its error (root-mean-square deviation) is above a threshold (we use 0.1 and 20 respectively), the candidate merger is abandoned. The intuition here is that, if two sequences belong to the same text line, the regression line fit to their combined fixations will have a gradient close to 0 and low regression error. Of the candidate mergers that remain, the pair of sequences with the lowest error are merged and added to the pool of sequences, replacing the original two sequences and reducing their number by one. This process is repeated until no further mergers are possible.

@@

The \texttt{regress} algorithm, which is closely based on \possessivecite{Cohen:2013} R package FixAlign, treats the fixations as a cloud of unordered points and fits $m$ regression lines to this cloud. These regression lines are parameterized by a slope, vertical offset, and standard deviation, and the best parameters are obtained by minimizing\footnote{In both the \texttt{regress} and \texttt{stretch} algorithms, we used the \texttt{minimize} function from the Python library SciPy \parencite{Virtanen:2020}. We have not systematically compared the choice of optimizer settings.} an objective function that determines the overall fit of the lines through the fixations. The algorithm has six free parameters which are used to specify the lower and upper bounds of the slope, offset, and standard deviation. Here we directly adopt FixAlign's defaults: $[-0.1, 0.1]$, $[-50, 50]$, and $[1, 20]$ respectively. Once the $m$ best-fitting regression lines are obtained, \texttt{regress} assigns each fixation to the highest-likelihood regression line, which itself is associated with a text line.

\NEWMATERIAL{\texttt{regress} tracks FixAlign very closely, except that we did not implement the ``run rule,'' an option that is switched on by default in FixAlign. This option maps ambiguous fixations to the same line as the surrounding fixations, if the surrounding fixations were classified unambiguously \parencite[p.~680]{Cohen:2013}. Cohen's run rule is a more general method that could in principle be applied to the output of any algorithm, so in the interest of isolating the core concept of FixAlign and comparing all algorithms on a level playing field, we did not to implement the option here.}

@@

As far as we know, the \texttt{split} algorithm takes an approach that is distinct from anything previously reported, although it is conceptually similar to \texttt{segment}. Like \texttt{segment}, the \texttt{split} algorithm works on the principle of splitting the fixation sequence into subsequences by first identifying the return sweeps. However, \texttt{split} is not restricted to finding exactly $m-1$ return sweeps; instead, it identifies the most likely set of return sweeps, however many that turns out to be. There are various ways of approaching this classification problem, but here we use \textit{k}-means clustering to partition the set of saccades into exactly two clusters. Since return sweeps are usually highly divergent from normal saccades \NEWMATERIAL{(i.e., a return sweep is usually represented by a large negative change on the \textit{x}-axis)}, one of the two clusters will invariably contain the return sweeps, which can then be used to split the fixation sequence into subsequences. However, since this is not guaranteed to produce $m-1$ return sweeps (and therefore $m$ subsequences), an order-based mapping is not possible, so \texttt{split} must use absolute position: Subsequences are mapped to the closest text lines in absolute terms. \texttt{split} has the advantage of generally finding all true return sweeps, and even if it identifies some false positives, the resulting subsequences can still be mapped to the appropriate lines by absolute position. However, this also means the algorithm is less resilient to vertical drift issues.

\subsection{Stretch}

\NEWMATERIAL{The \texttt{stretch} algorithm is loosely based on the the method proposed by \textcite{Lohmeier:2015} and shares some similarities with \textcite{MartinezGomez:2012} and \textcite{Nussli:2011}. \possessivecite{Lohmeier:2015} original method was designed for the reading of source code and therefore takes advantage of the fact that code has very irregular line lengths and indentation levels. The method works by finding an \textit{x}-offset, \textit{y}-offset, and scaling factor that, once applied to the fixations, minimizes alignment error between the fixations and lines of text.}

\NEWMATERIAL{The framework we adopt herein never adjusts the \textit{x}~values, and we also assume that an ordinary passage of text is being read, so line length is substantially more constant than during code reading and therefore less informative. Therefore, we simplified the original method by dispensing with all dependencies on the \textit{x}~values. Instead, \texttt{stretch} finds a \textit{y}-offset, $o^\ast$, and a vertical scaling factor, $s^\ast$, that minimizes the sum absolute difference between the corrected fixation positions ($f_y s + o$) and the corrected fixation positions once attached to their closest lines. The equations presented in \textcite[pp.~37--38]{Lohmeier:2015} therefore simplify to:}

	\begin{equation}
	o^\ast, s^\ast = \operatornamewithlimits{arg\,min}_{o,s} \sum_{f \in F} | (f_y s + o) - \mathrm{attach}(f_y s + o) |,
	\label{equation1}
	\end{equation}

\noindent
\NEWMATERIAL{where $\mathrm{attach}(\cdot)$ returns the \textit{y}-axis position of the nearest line of text. In other words, the algorithm seeks a transformation of the fixations that results in minimal change following the application of \texttt{attach}.}

\NEWMATERIAL{To constrain the minimization problem, the user must specify appropriate lower and upper bounds for the offset and scaling factor, resulting in four free parameters. Here we adopt offset bounds of $[-50, 50]$, following the \texttt{regress} algorithm, and scaling factor bounds of $[0.9, 1.1]$. Effectively, this means the algorithm can move the set of fixations up or down by up~to 50~pixels and stretch their positions on the vertical axis by between 90\% and 110\%. While approaching the problem from a different angle, \texttt{stretch} is computationally similar to \texttt{regress}, except that it emphasizes systematic offset issues rather than systematic slope issues.}

@@

The final algorithm we consider, \texttt{warp}, is novel to this paper, although it is mostly a wrapper around a preexisting algorithm---dynamic time warping \parencite[DTW;][]{Vintsyuk:1968,Sakoe:1978}. DTW was used by the \texttt{compare} algorithm to provide a measure of dissimilarity between a gaze line and a text line. To our knowledge, however, there have been no previous reports of DTW being used directly to align fixations to text lines. This is somewhat surprising because DTW is the natural computational choice for tackling drift and alignment problems. The closest previously-described method is \textcite{Carl:2013}, who uses a basket of reading-related measures to place a cost on different paths through a lattice of fixation-to-character mappings and selects the path with minimal cost. This is quite complex, however, and we consider it to be a special case of \texttt{warp}, which is a direct application of the standard DTW algorithm to eye tracking data.

\NEWMATERIAL{DTW is typically useful when you have two sequences, not necessarily of the same length, and you want to (a) calculate how similar they are (as is the case in the \texttt{compare} algorithm) or (b) align the two sequences by mapping each element in one sequence to a corresponding element in the other. For example, DTW may be used to calculate the similarity between a signature, which can be expressed as a sequence of \textit{xy}-coordinates over time, and a reference signature \parencite[e.g.,][]{Lei:2005,Riesen:2018}. Importantly, the two sequences do not need to be perfectly matched in terms of overall magnitude or patterns of acceleration and deceleration for a good alignment to be found. In the case of signature verification, for example, it does not matter if the candidate signature has the same size as the reference or that it was drawn at the same speed, what matters is that there is a good match in the overall shape and that the strokes were drawn in the same order. DTW finds many other applications in, for example, genomics \parencite{Aach:2001}, medicine \parencite{Caiani:1998}, and robotics \parencite{Vakanski:2014}.}

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/fig02_single_column.eps}}
	\vspace*{2pt}
	\caption{Illustration of the \texttt{warp} algorithm. The veridical fixation sequence is represented in blue, and the expected fixation sequence (the sequence of word centers) is represented in red. The dashed black lines show the DTW warping path---the optimal way to align the two sequences, such that the sum of the Euclidean distances between matched points (i.e., the sum of the dashed lines) is minimized.}
	\label{fig02}
	\end{figure}
